---
title: "Project 2 : WSOP Main Event data analysis"
format: pdf
author: "Group Members: Pyo & Frank"
---


```{r setup}
#| message: false
#| warning: false
#| include: false


library(tidyverse)
library(DBI)
library(duckdb)

```

# Workflow & Reproducibility Guide

To ensure this entire workflow is fully reproducible, follow these steps:

Download the raw dataset from Kaggle (link provided in the Sources section).

Place the file WSOP__1971-2025.csv in your working directory.

Run this Quarto document. The R setup will automatically clean the data, normalize it into 3rd Normal Form (3NF) relational tables, export them as temporary CSVs, and ingest them into a local DuckDB file (poker_analysis.duckdb).

### ---------------------------------------------------------------------------
# Accessing data & Designing the database

```{r}

wsop_old <- read_csv("WSOP__1971-2025.csv")

wsop_raw <- wsop_old %>%
  mutate(
    country = case_when(
      country %in% c("US", "USA") ~ "USA",
      is.na(country) ~ "Unknown",
      TRUE ~ country
    ),
    
    city = replace_na(city, "Unknown"),
    state = replace_na(state, "Unknown")
  )

# Players

players <- wsop_raw %>%
  select(name, city, state, country) %>%
  distinct() %>%
  mutate(player_id = row_number()) %>%
  relocate(player_id) 

# Tournaments
tournaments <- wsop_raw %>%
  select(year, entries, prizepool, buyin, start, finish) %>%
  distinct() %>%
  mutate(event_id = row_number()) %>%
  relocate(event_id) 

# Results
results <- wsop_raw %>%
  left_join(players, by = c("name", "city", "state", "country")) %>%
  left_join(tournaments, by = c("year", "entries", "prizepool", "buyin", "start", "finish")) %>%
  select(player_id, event_id, place, prize) %>%
  mutate(result_id = row_number()) %>%
  relocate(result_id)  


write_csv(players, "players.csv")
write_csv(tournaments, "tournaments.csv")
write_csv(results, "results.csv")

# DuckDB
con <- dbConnect(duckdb(), dbdir = "poker_analysis.duckdb")

# --- Players ---
dbExecute(con, "DROP TABLE IF EXISTS players;")
dbExecute(con, "CREATE TABLE players (
    player_id INTEGER PRIMARY KEY,
    name VARCHAR(255),
    city VARCHAR(255),
    state VARCHAR(50),
    country VARCHAR(100)
);")
dbExecute(con, "COPY players FROM 'players.csv' (FORMAT CSV, HEADER);")

# --- Tournaments ---
dbExecute(con, "DROP TABLE IF EXISTS tournaments;")
dbExecute(con, "CREATE TABLE tournaments (
    event_id INTEGER PRIMARY KEY,
    year INTEGER,
    entries INTEGER,
    prizepool BIGINT,
    buyin INTEGER,
    start DATE,
    finish DATE
);")
dbExecute(con, "COPY tournaments FROM 'tournaments.csv' (FORMAT CSV, HEADER);")

# --- Results ---
dbExecute(con, "DROP TABLE IF EXISTS results;")
dbExecute(con, "CREATE TABLE results (
    result_id INTEGER PRIMARY KEY,
    player_id INTEGER,
    event_id INTEGER,
    place INTEGER,
    prize BIGINT
);")
dbExecute(con, "COPY results FROM 'results.csv' (FORMAT CSV, HEADER);")

dbGetQuery(con, "SELECT * FROM players LIMIT 5;")
```





### ---------------------------------------------------------------------------
# Querying data using SQL

# Global Dominance: Top 10 Countries by Total WSOP Winnings
# Analyzing which countries have produced the highest cumulative prize money through successful tournament finishes.

```{sql connection=con}
SELECT 
    p.country, 
    SUM(r.prize) AS total_winnings,
    COUNT(r.result_id) AS total_cashes
FROM players AS p
JOIN results AS r ON p.player_id = r.player_id
WHERE p.country IS NOT NULL
GROUP BY p.country
ORDER BY total_winnings DESC
LIMIT 10;
```
#Top 10 Career Earners in Main Event History
#Identifying the individual players who have secured the most prize money across their entire WSOP careers. 
```{sql connection=con}
SELECT 
    p.name, 
    SUM(r.prize) AS career_earnings
FROM players AS p
JOIN results AS r ON p.player_id = r.player_id
GROUP BY p.name
ORDER BY career_earnings DESC
LIMIT 10;
```
#Top 10 Highest ROI Performances
# Calculating the Return on Investment (ROI) to determine which tournament runs were the most profitable relative to the buy-in.
```{sql connection=con}
SELECT 
    t.year, 
    t.buyin, 
    r.prize,
    ((r.prize - t.buyin) * 100.0 / t.buyin) AS roi_percentage
FROM tournaments AS t
JOIN results AS r ON t.event_id = r.event_id
WHERE r.prize > 0
ORDER BY roi_percentage DESC
LIMIT 10;
```

# Most Reliable Players with 10+ Career Cashes
# Filtering for elite players who consistently finish in the money, demonstrating long-term skill and stability.
```{sql connection=con}
SELECT 
    p.name, 
    COUNT(r.result_id) AS cash_count,
    AVG(r.place) AS average_place
FROM players AS p
JOIN results AS r ON p.player_id = r.player_id
GROUP BY p.name
HAVING COUNT(r.result_id) >= 10
ORDER BY average_place ASC
LIMIT 10;
```

# Year-over-year growth in entries
```{sql connection=con}
SELECT
    year,
    SUM(entries) AS total_entries,
    LAG(SUM(entries)) OVER (ORDER BY year) AS previous_year_entries,
    ROUND(
        100.0 * (SUM(entries) - LAG(SUM(entries)) OVER (ORDER BY year))
        / LAG(SUM(entries)) OVER (ORDER BY year), 2
    ) AS yoy_entry_growth_pct
FROM tournaments
GROUP BY year
ORDER BY year;
```

# Top 10 years by total prizepool
```{sql connection=con}
SELECT
    year,
    SUM(prizepool) AS total_prizepool,
    SUM(entries) AS total_entries,
    COUNT(*) AS total_events
FROM tournaments
GROUP BY year
ORDER BY total_prizepool DESC
LIMIT 10;
```

# Buy-in distribution by year
```{sql connection=con}
SELECT
    CASE
        WHEN buyin < 1000 THEN 'Under $1K'
        WHEN buyin BETWEEN 1000 AND 4999 THEN '$1K-$4,999'
        WHEN buyin BETWEEN 5000 AND 9999 THEN '$5K-$9,999'
        ELSE '$10K+'
    END AS buyin_tier,
    COUNT(*) AS number_of_tournaments,
    AVG(entries) AS avg_entries,
    AVG(prizepool) AS avg_prizepool
FROM tournaments
GROUP BY buyin_tier
ORDER BY number_of_tournaments DESC;
```

# Most Profitable Buy-In Levels
```{sql connection=con}
SELECT
    t.buyin,
    COUNT(r.result_id) AS total_cash_results,
    ROUND(AVG(r.prize), 2) AS avg_prize,
    MAX(r.prize) AS max_prize,
    SUM(r.prize) AS total_prize_awarded
FROM tournaments AS t
JOIN results AS r
    ON t.event_id = r.event_id
WHERE t.buyin IS NOT NULL
GROUP BY t.buyin
HAVING COUNT(r.result_id) >= 20
ORDER BY avg_prize DESC
LIMIT 10;
```


### ---------------------------------------------------------------------------
# Creating data visualizations
```{r}
country_data <- dbGetQuery(con, "
    SELECT p.country, SUM(r.prize) as total 
    FROM players AS p 
    JOIN results AS r ON p.player_id = r.player_id 
    WHERE p.country IS NOT NULL
    GROUP BY p.country 
    ORDER BY total DESC 
    LIMIT 5")

ggplot(country_data, aes(x = reorder(country, total), y = total, fill = country)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::dollar_format()) +
  coord_flip() +  
  labs(
    title = "Top 5 Poker Powerhouses by Total Winnings",
    subtitle = "Analysis of WSOP Main Event Data (1971-2025)",
    x = "Country",
    y = "Total Prize Money ($)"
  ) +
  theme_minimal()

```

# YoY winning table
```{r}
yoy_growth <- dbGetQuery(con, "
SELECT
    year,
    SUM(entries) AS total_entries
FROM tournaments
GROUP BY year
ORDER BY year
")

ggplot(yoy_growth, aes(x = year, y = total_entries)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "WSOP Tournament Participation Over Time",
    subtitle = "Total yearly entries across all tournaments",
    x = "Year",
    y = "Total Entries"
  ) +
  theme_minimal()
```


```{r}
# terminate the SQL connection

DBI::dbDisconnect(con, shutdown = TRUE)

```

### ---------------------------------------------------------------------------

# Sources

* https://www.kaggle.com/datasets/cviaxmiwnptr/wsop-main-event-results-1971-2024?resource=download

* web link/document title/LLM prompt


### ---------------------------------------------------------------------------
